Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient

### 2022-12-08 manuscript/manuscript.Rmd
update references
discussion changes

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly use 1) geochemical proxy data, 2) ecological or lithological climate indicators, or 3) earth system modelling.

Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009]. To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g. @Veizer2015]. This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022]. Ecological and lithological climate data provide an alternative method of reconstructing palaeotemperatures, using e.g. leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021]. Despite some recent advances [@Chandra2021], ecological and lithological climate indicators are still rarely used in quantitative palaeoclimate reconstructions, particularly for the marine realm. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???]. However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.

Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of workers. Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].

@@

**Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques. Parametric methods include polynomials [@Greenwood1995b; @Bij2009] and a cosine function [@Inglis2020]. Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig. 2). Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets. Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists. The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).

>>>>

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.

Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009]. Palaeobiological and lithological climate data provide a different method of reconstructing palaeotemperatures, using e.g. leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021]. To capture complex dependencies between proxies and climate data, machine learning approaches [@Chandra2021] or articifical neural networks [@Malmgren1997; @Lauchstedt2017] are sometimes used. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???]. However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.

To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings. Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g. @Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022]. Reconstructions that take into account the spatial distribution of climate data are... ### HERE###
provide an alternative that can be employed more readily by a broad range of workers. Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].

@@

**Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques. Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020]. Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig. 2). Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets. Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists. The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods). A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies. An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
